---
title: "Introducing the finbif R package"
author: "Will"
date: 2021-09-01
categories: ["R"]
tags: ["R", "FinBIF"]
---

```{r setup, eval = TRUE, echo = FALSE, cache = FALSE}
library(details)
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  details.summary = "Click to show/hide output.",
  details.tooltip = NULL,
  details.comment = "#>"
)

library(finbif)
options(finbif_tz = "Europe/Helsinki", width = 100)
```
The `{finbif}` R package makes the publicly available data in FinBIF easily
accessible to programmers. Biodiversity information is available on taxonomy
and taxon occurrence. Occurrence data can be filtered by taxon, time, location
and other variables. The data accessed are conveniently preformatted for
subsequent analyses.

## Installing the finbif package
You can install the current stable version of `{finbif}` from
[CRAN](https://cran.r-project.org),
```{r install_cran, eval = FALSE, cache = FALSE}
install.packages("finbif")
```

You can also install the latest development version of `{finbif}` from
[GitHub](https://github.com),
```{r install_gh, eval = FALSE, cache = FALSE}
remotes::install_github("luomus/finbif@dev")
```

## Loading the finbif package
```{r library, eval = FALSE, cache = FALSE}
library(finbif)
```

## Getting a FinBIF access token
To use the FinBIF API you must first request and set a personal access token.
You can request an API token to be sent to your email address with the function
`finbif_get_token()`.
```{r request-token, eval = FALSE, cache = FALSE}
finbif_request_token("your@email.com")
```

Copy the access token that was sent to your email and set it as the environment
variable `FINBIF_ACCESS_TOKEN` either for the current session,
```{r set-token, eval = FALSE, cache = FALSE}
Sys.setenv(
  FINBIF_ACCESS_TOKEN = "xtmSOIxjPwq0pOMB1WvcZgFLU9QBklauOlonWl8K5oaLIx8RniJLrvcJU4v9H7Et"
)
# Note: the above is not a real access token. Do not try using it.
```
, or by adding it to a `Renviron` startup file (see
[here](https://rviews.rstudio.com/2017/04/19/r-for-enterprise-understanding-r-s-startup/)
for details).

## Getting occurrence data
You can download occurrence data from the FinBIF database as a `data.frame` with
the `finbif_occurrence()` function.
```{r occurrence}
finbif_occurrence("Cygnus cygnus", n = 100)
```

You can search for multiple taxa at once and filter the records with the
`filter` argument.
```{details occurrence2}
finbif_occurrence(
  "Cygnus cygnus", 
  "Cygnus olor",
  filter = list(coordinates_uncertainty_max = 100)
)
```

## Plotting occurrence data
The `{finbif}` package has a number of inbuilt functions for plotting (see e.g.,
`breaks_xy()` and `hist_xy()`). There is also an inbuilt dataset that can be
used to plot the border of Finland (`?finland_map`). Together these utilities
can be used to plot occurrences after they have been downloaded from FinBIF. For
example, the following can be used to plot the density of Eurasian Jay
occurrences from Finland.

<details closed>
<summary><span title=''>Click to show/hide code.</span></summary>
```{r plot-occurrences, eval = FALSE, cache = FALSE}
# Download all the occurrences of Eurasian Jay in Finland
# that have coordinates accurate to at least 100m
jays <- finbif_occurrence(
  taxa   = "Eurasian Jay",
  filter = c(
    coordinates_uncertainty_max = 100,
    country                     = "Finland"
  ),
  n      = 2e4,
  quiet  = TRUE
)

# Compute the density of occurrences in 1/4 degree cells and plot as a heatmap
with(
  data = c(jays, finland_map),
  expr = {
    par(mar = c(5, 5, 1, 1), las = 1)
    # compute a 2d histogram from the occurrences
    breaks  <- breaks_xy(bbox, .25) # breakpoints every 1/4 of a degree
    density <- hist_xy(xy = list(lon_wgs84, lat_wgs84), breaks)
    # plot the histogram as a heatmap
    image(density,
          asp    = 2.4,
          breaks = 2^seq(0, 12), # breakpoints for the gridcell colours
          col    = hcl.colors(12, rev = TRUE),
          xlab   = "Longitude",
          ylab   = "Latitude",
          panel.first = grid())
    legend("topright",
           inset  = c(0, .01),
           legend = expression(2^12, "", "", 2^6, "", "", 2^0),
           fill   = hcl.colors(7),
           border = NA,
           bty    = "n",
           adj    = c(0, 0.25), 
           x.intersp = .2,
           y.intersp = .5)
    # add the Finnish border
    polygon(x = vertices, lwd = .2)
  }
)
```
</details>
```{r plot-occurrences-eval, echo = FALSE, fig.cap = "", fig.width = 5, fig.height = 7}
# Download all the occurrences of Eurasian Jay in Finland
# that have coordinates accurate to at least 100m
jays <- finbif_occurrence(
  taxa   = "Eurasian Jay",
  filter = c(
    coordinates_uncertainty_max = 100,
    country                     = "Finland"
  ),
  n      = 2e4,
  quiet  = TRUE
)

# Compute the density of occurrences in 1/4 degree cells and plot as a heatmap
with(
  data = c(jays, finland_map),
  expr = {
    par(mar = c(5, 5, 1, 1), las = 1)
    # compute a 2d histogram from the occurrences
    breaks  <- breaks_xy(bbox, .25) # breakpoints every 1/4 of a degree
    density <- hist_xy(xy = list(lon_wgs84, lat_wgs84), breaks)
    # plot the histogram as a heatmap
    image(density,
          asp    = 2.4,
          breaks = 2^seq(0, 12), # breakpoints for the gridcell colours
          col    = hcl.colors(12, rev = TRUE),
          xlab   = "Longitude",
          ylab   = "Latitude",
          panel.first = grid())
    legend("topright",
           inset  = c(0, .01),
           legend = expression(2^12, "", "", 2^6, "", "", 2^0),
           fill   = hcl.colors(7),
           border = NA,
           bty    = "n",
           adj    = c(0, 0.25), 
           x.intersp = .2,
           y.intersp = .5)
    # add the Finnish border
    polygon(x = vertices, lwd = .2)
  }
)
```
